Phylotyper: {\it In silico} predictor of subtypes from gene subtypes

### 2017-05-30 manuscript/response_to_reviewers.txt

Although the GitHub page is well laid out and easy enough to follow, please consider including some sort of example to follow along with. As a newcomer to the application it took me a little while to figure out how to get everything exactly right for an initial run. I think a very small example to follow and run that is located early in the GitHub instructions would really help to just get things going.

*Response:
We agree with the reviewer's suggestion that a working example for new users will enhance the adoption and usability of Phylotyper.  We have added an "Example" section to the README.md file (which is also the landing page for Phylotyper's GitHub repository: https://github.com/superphy/insilico-subtyping/#examples) along with example data that demonstrates how to 1) predict subtypes for schemes already packaged in Phylotyper and 2) how to add and run user-defined subtype schemes.

There lies a slight issue in the "user generated sub-type schemes" in that the results that you get with Phylotyper are highly dependant on the quality of the sub-type scheme provided. Since this tool will likely be used in outbreak situations and analyses, do you have any plans to address the quality of the sub-type scheme that was used? Do you plan on having a section in the GitHub page that allows users to send new sub-type schemes for validation and posting?

*Response*:
The reviewer is correct that Phylotyper's performance is dependent on the input reference sequences used to build the subtype predictor. To mitigate poor prediction from unsuitable or inaccurate inputs, we included in the build pipeline for the local user-submitted subtypes, a series of checks to automatically test the performance of any submitted subtype scheme. These checks are now described in the Implementation section.

While these checks identify major breakdowns in the training set used by Phylotyper, we note there is no check to identify missing data, specifically, missing subtypes or key subtypes gene sequences that define the range of the subtype sequence space. For any current general subtype predictor (i.e. not predictors that replicate PCR in silico, for instance), performance is dependent on the reference input set. In dealing with missing data, we demonstrate that Phylotyper is significantly more robust than a direct sequence similarity-based predictor. It is less likely to generate a type-I error when encountering a subtype not found in the training set. The Phylotyper method estimates the uncertainty associated with a subtype prediction. When test sequences that are phylogenetically distant from known reference sequences or in a region of the phylogenetic tree that is rapidly evolving, the Phylotyper score will reflect the degree of uncertaintity. Most often the sequence for a novel or missing subtype will be relatively phylogenetically distant from the other subtype sequences in the tree and its Phylotyper uncertaintity score will reflect this. Only in cases where the novel/missing subtype is closely phylogenetically related to an existing subtype in the training set, will Phylotyper generate a false positive prediction. This property of Phylotyper makes it more robust when working with incomplete training data.

To enhance the users understanding of the checks that are provided in the build feature and the potential blindspots in creating their own subtype predictors with Phylotyper, we have greatly expanded the methods section and the README provided with the software. The added content describes the two checks, how they work and what they detect. In manuscript we also provide guidelines about the inputs needed for an effective subtype predictor. We also included in the manuscript a new analysis that demonstrates how Phylotyper versus a BLAST-based predictor performs when an entire subtype class is removed from the trainging set and used as test input. This analysis and results have been added to the Results and Discussions sections.

The reviewer suggests having a means for users to submit subtypes for evaluation and posting to a shared repository. We agree that this would be a helpful system. It will allow microbiologist specialists to submit and improve subtype schemes and increase coverage. We mention in the manscript and on the README, that users can contact us and submit new or modified subtype schemes. We will check the submitted schemes and add them to our Github repository for general use.

@@

3.      Please provide more details regarding the “assumptions of the Phylotyper approach” and how/why they would be violated when building new subtype schemes.

*Response*: 
The Phylotyper assumptions are now outlined in the Implementation section. The revised Implementation section covers the basis of the assumptions, how they can be violated, and two built-in mechanisms we developed to alert users of possible violations.

Results:
4.      Whiteside et al compare here their tool with another BLAST based approach that they have developed for the comparison, and not methods published by other groups. The description of the method used for both the BLAST method and the cross validation is imprecise and would benefit from more being more detailed within the main article rather than in the supplementary material. Moreover, a comparison - at least theoretical - with other existing tools would strengthen the uniqueness of Phylotyper highlighted in the discussion.

*Response*:
Our main goal in the performance assessment was to compare how the prediction methods generalize to an independent dataset. There may exists differences in the reference training sets that underpin the tools, and these differences may mask the true performance of the method and we wanted to evalute the predictive capability of the methods themselves. SeroTypeFinder is provided as a web service. The software for SeroTypeFinder has not been made publicly available. Without access to the software, we cannot manipulate the training and test datasets to perform cross-validation analysis. The EcOH tool works directly on unassembled small reads and not assembled genome sequences. A direct comparison of the performance with these tools with Phylotyper is not possible. The intention of using a general BLAST-based approach is demonstrate the benefit of Hidden State Prediction with a phylogenetic tree over approaches that directly match input sequences against a reference database. BLAST, as used in our assessment with a percent identity and coverage cutoff, is the basis of the SeroTypeFinder tool. EcOH employs a more sophisticated algorithm for matching alleles in order to filter out spurious small reads that might align by chance. At its core, it is still matching alleles and does not develop a model of subtype/allele sequence evolution to define boundaries on the subtypes. We feel that our performance comparison to the generalized BLAST-based method does make a valid case for using HSP in subtype prediction. We did add a direct comparison of Phylotyper to Vtyper for Stx genes. The performance results are summerized in the Results section.

To address the reviwers comment that the BLAST and cross-validation method descriptions are imprecise, we have added more thorough descriptions of the BLAST method and the cross-validation analysis in the manuscript. Our aim with the revised descriptions is to provide readers sufficient information give them a complete understanding of the mechanics of the methods. Implementation details, such as percent identity cutoff or number of iterations, remain in Supplementary Information. This was done to fit with page constraints of the Application Note format.

@@

As suggested, we also added a theoretical comparison of the tools to the Discussion; the theoretical comparison looks at use cases, potential issues, and additional benefits outside of performance for the tools.

Discussion:
5.      There is no data to support the claim that Phylotyper easily identifies novel subtype. Did the author try to remove all alleles of one subtype and compute the predictions to see if/how phylotyper creates a consistent subtype? Also, when a new subtype is identified, does Phylotyper includes it in his typing scheme automatically? What would be the recommended tests/further analysis recommended to researchers using Phylotyper in the case a new subtype is identified?

*Response*:
The point we were trying make was that more frequently, Phylotyper correctly handles novel subtypes than existing methods. In our emperical testing we found Phylotyper makes fewer Type-I errors when encountering a novel allele of an existing subtype or a allele with a entirely new subtype not found in reference training set. In former case, Phylotyper returns non-significant/undetermined. We have clarified this point throughout the manuscript, removing references to Phylotyper identifying novel subtypes and replacing it with the more nuanced description that Phylotyper more frequently properly classifies novel subtypes as a negative result or correctly classifies novel alleles with subtypes in the database. To strengthen this statement, we have as the reviewers suggest, remove entire subtypes from the training set and calculated the false positive rate. Phylotyper was found to have a much lower false positive rate when classifying genes with unknown subtypes.

Phylotyper does not have the ability to create subtype schemes de novo, or to add new subtype classes on the fly. We have not explored this potential application of the Phylotyper method. It would be an interesting avenue to explore as using a phylogenetic model of subtype evolution to define boundaries for new or existing subtypes might make a nice complement for existing de novo stragies like MLST that assign every new allele a separate subtype. However, our goal in this manuscript is to present an alternative genome-based subtype prediction method. De novo subtype creation is outside our intended applications for Phylotyper. Again, we corrected all references to Phylotyper identifing novel subtypes.

@@

6.      “Subtypes are mainly used [..] the main uses of subtype information”. Please revise this paragraph, to allow a better logical flow. Also, Debroy et al have shown the discrepancy between between serotype-based typing and the genetic similarity of O antigen gene clusters, suggesting to replace serotyping by molecular typing. They did not particularly emphasize discrepancies between existing molecular typing scheme and whole genome data as suggested by Whiteside et al.

*Response*:

We have revised this paragraph and clarified our point on the suitability of Phylotyper for predicting subtypes from WGS; namely that Phylotyper helps identify discrepancies between a subtype and genome region that is used for prediction. We cite Debroy et al., as an example of how the associated genetic data and subtypes can have discrepancies.

@@

*Response:
Complete or nearly complete loci sequences are needed for correctly placing the loci in the phylogenetic tree. To prevent incomplete loci being used, we set coverage cutoffs of 90% in the BLAST detection step. No alleles with less than this coverage to one of the reference set sequences will pass the detection step. Users are notified when no loci I have added a description of the requirement for complete loci sequences and the checks we have built into Phylotyper to the Discussion.

2) Please clarify how novel subtypes/alleles are identified and reported.

*Response
Phylotyper is significantly more robust than a direct sequence similarity-based predictor in handling novel subtypes or alleles. We do not mean that Phylotyper performs unsupervised construction of new subtype schemes (although the underlying method could be adapted for this task), rather, that Phylotyper has significantly lower false positive rates when performing prediction on an allele not in the reference set or an allele for which there is no example with the same subtype in the reference set (i.e. novel subtype). In the case of a novel subtype, Phylotyper would not assign the allele any subtype.  Compare this result with a sequence similarity approach, which, in most cases assign the novel subtype the closest subtype based on sequence.

To clarify this distinction, we have added and modified text.  We have also performed and included a subsequent analysis comparing the accuracy of the BLAST-based approach to the Phylotyper approach when the entire subtype group is withheld from the training set to simulate performance when encountering a novel subtype. This analysis shows that Phylotyper is a marked improvement over the sequence-similarity approach, and more frequently returns a negative result.

@@

3) As there is a trend moving to core-genome or whole-genome MLST typing schemes, please comment if Phylotyper could be scaled up to handle that. How would horizontal gene transfer or recombination affect the Phylotyper results?

*Response:
The Phylotyper approach is not likely not applicable to core-genome or whole-genome MLST schemes. The value that Phylotyper provides is that it can take as input gene sequences and return predictions for subtypes based on phenotype or subtypes that cannot be directly inferred from the gene sequences alone. For example, PCR or restriction digestion-based subtyping schemes are an abstraction of the underlying gene sequences and in most cases, cannot be directly inferred from an in silico genome sequence. We have demonstrated the ability of Phylotyper to accurately predict a PCR-based subtype (Stx1 and Stx2) as well as a phenotype-based scheme (O-type:H-type serotype).

Subtypes that are directly identified from the genome sequence do not require Phylotyper's additional prediction step. In MLST, the combination each unique allele sequence defines the type.  Potentially, and this is perhaps what the reviewer is refering to, a specific region of the genome that is smaller than the sequence regions used in an MLST, exists that can be used as a proxy to predict MLST subtype. The benefit in this case, is that the entire genome would not need to be sequenced (or multiple alleles interogated) in order to predict MLST type. There is precedent for this type of approach.  The spa and coa genes in Staphylococcus aureus are used as gene markers to infer phylogenetic and population variation.  The spa and coa types have noted concordance with MLST typing. While this is in interesting potential new application for Phylotyper, it is outside the scope of what we originally designed Phylotyper for.

Horizontal gene transfer or recombination would only impact the result of Phylotyper in situations where the genotype is only correlated with subtype and not causative. If the sequences used to build the subtype scheme produce gene products that determine the subtype, such as the O-antigen processing genes for E. coli being used to predict O-group serotypes, then the prediction would directly be linked to underlying genome. If however unrelated, correlated sequence is used, HGT could disconnect a particular genotype from its subtype. We have added to the manuscript recommendations for users building subtype schemes to focus on genomic regions that have a direct biological role in determining the subtype to reduce possibility of impact by HGT.

Minor:
1) Typo in line 49: also not aslo

*Response*:
Fixed

2) Line 51 mentioned assumptions for building new subtype pipeline.  It'll be good to explain that or if space is an issue, point to the portion of the manual that explains these assumptions

*Response*:
We have added descriptions of these assumputions to the mansucript.

### 2017-05-30 manuscript/phylotyper_revised.tex
EDIT: finished response to reviewers

\abstract{\textbf{Summary:} Whole genome sequencing (WGS) is being adopted in public health for improved surveillance and outbreak analysis.
Subtyping; a  is used in public health to infer phenotypes and distinguish bacterial strain groups.
\textit{In silico} tools that predict molecular subtypes from sequences data are needed to transition historical data to WGS-based protocols.
Phylotyper is a novel solution for \textit{in silico} subtype prediction from gene sequences.
Designed for incorporation into WGS pipelines, it is a general prediction tool that can be applied to most molecular subtype schemes.
Phylotyper uses phylogeny to model the evolution of the subtype and infer subtypes for unannotated sequences.
The phylogenic framework in Phylotyper improves accuracy, provides useful contextual feedback, and is more capable of identifying novel subtypes over approaches based solely on sequence similarity.\\
\textbf{Availability and Implementation:} Phylotyper is a python and R package. It is available from: \url{https://github.com/superphy/insilico-subtyping}.\\
\textbf{Contact:} \href{matthew.whiteside@phac-aspc.gc.ca}{matthew.whiteside@phac-aspc.gc.ca}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

Currently, the Phylotyper package includes the following subtype schemes for \emph{Escherichia coli} (\textit{E. coli}): Stx, intimin and serotype O- and H-types  (Supplementary Table~1).
However, the Phylotyper software also has the capability to add new subtype schemes (instructions are provided here: \url{https://github.com/superphy/insilico-subtyping}). 
Creating a new subtype scheme will save the required reference files, allowing newly added schemes to be easily re-run from Phylotyper.
To add a new subtype scheme for use in Phylotyper, users require a training set of homologous sequences with assigned subtype whose phylogenetic grouping is predictive of the subtype.
Phylotyper assumes that the provided training sequences are 1) homologous, specifically, they are suitable for alignment and phylogenetic reconstruction 2) the sequence phylogeny is correlated with the subtype distribution and 3) the set is representative of the range of sequences that make up a subtype.
Where possible, checks are built-in to the \texttt{new} pipeline to validate the submitted reference set.
Specifically, each new subtype is subject to two tests in the \texttt{new} pipeline and results are reported to the user.
The first test checks that the distribution of inter-patristic phylogenetic distances between the instances of the same subtype is both smaller and distinct from non-identical subtypes.
This test can identify isolated cases of potentially mislabelled subtype genes that are tightly clustered with other subtypes.
A second check computes the accuracy measure, F-score, through a leave-one-out cross-validation; a procedure which uses each sequence in the training set as a test input to estimate positive and negative prediction rates (Supplementary Methods).
The performance metrics: recall and F-score rapidly decrease as the correlation between the training set phylogeny and subtype distribution decreases (Supplementary Figure)
These checks will verify that the training set sequences' phylogenetic grouping is predictive of the subtype.
There is no check that can confirm the training set covers all subtypes in a particular scheme; it is important that the user monitor and update schemes as gaps are identified.
Phylotyper is designed to return a \texttt{non-significant/undetermined} result when encountering a unknown subtype that has no representative in the training set.
Phylotyper is comparatively more robust than other methods as it generates fewer Type-I errors when encountering novel alleles or subtypes not present in the training set (see Results section for details).

>>>>

\abstract{\textbf{Summary:} Whole genome sequencing (WGS) is being adopted in public health for improved surveillance and outbreak analysis.
Subtyping; a  is used in public health to infer phenotypes and distinguish bacterial strain groups.
\textit{In silico} tools that predict molecular subtypes from sequences data are needed to transition historical data to WGS-based protocols.
Phylotyper is a novel solution for \textit{in silico} subtype prediction from gene sequences.
Designed for incorporation into WGS pipelines, it is a general prediction tool that can be applied to most molecular subtype schemes.
Phylotyper uses phylogeny to model the evolution of the subtype and infer subtypes for unannotated sequences.
The phylogenic framework in Phylotyper improves accuracy and provides useful contextual feedback, then approaches based solely on sequence similarity.\\
\textbf{Availability and Implementation:} Phylotyper is a python and R package. It is available from: \url{https://github.com/superphy/insilico-subtyping}.\\
\textbf{Contact:} \href{matthew.whiteside@phac-aspc.gc.ca}{matthew.whiteside@phac-aspc.gc.ca}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

Currently, the Phylotyper package includes the following subtype schemes for \emph{Escherichia coli} (\textit{E. coli}): Stx, intimin and serotype O- and H-types  (Supplementary Table~1).
However, the Phylotyper software also has the capability to add new subtype schemes (instructions are provided here: \url{https://github.com/superphy/insilico-subtyping}). 
Creating a new subtype scheme will save the required reference files, allowing newly added schemes to be easily re-run from Phylotyper.
To add a new subtype scheme for use in Phylotyper, users require a training set of homologous sequences with assigned subtype whose phylogenetic grouping is predictive of the subtype.
Phylotyper assumes that the provided training sequences are 1) homologous, specifically, they are suitable for alignment and phylogenetic reconstruction 2) the sequence phylogeny is correlated with the subtype distribution and 3) the set is representative of the range of sequences that make up a subtype.
Where possible, checks are built-in to the \texttt{new} pipeline to validate the submitted reference set.
Specifically, each new subtype is subject to two tests in the \texttt{new} pipeline and results are reported to the user.
The first test checks that the distribution of inter-patristic phylogenetic distances between the instances of the same subtype is both smaller and distinct from non-identical subtypes.
This test can identify isolated cases of potentially mislabelled subtype genes that are tightly clustered with other subtypes.
A second check computes the accuracy measure, F-score, through a leave-one-out cross-validation; a procedure which uses each sequence in the training set as a test input to estimate positive and negative prediction rates (Supplementary Methods).
The performance metrics: recall and F-score rapidly decrease as the correlation between the training set phylogeny and subtype distribution decreases (Supplementary Figure)
These checks will verify that the training set sequences' phylogenetic grouping is predictive of the subtype.
There is no check that can confirm the training set covers all subtypes in a particular scheme; Phylotyper can only predict subtypes that are covered in the trainging set.
It is important that the user monitor and update schemes as gaps are identified.
Phylotyper is designed to return a \texttt{non-significant/undetermined} result when encountering a unknown subtype that has no representative in the training set.
Phylotyper is comparatively more robust than other methods as it generates fewer Type-I errors when encountering novel alleles or subtypes not present in the training set (see Results section for details).
